---
title: "Gap analysis workflow for multiple species"
output: 
  html_document:
vignette: >
  %\VignetteIndexEntry{Gap Analysis Workflow for Multiple Species}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


##

This workflow demonstrates how to import your own point data (occurrences) and SDMs (rasters), check all inputs, and run the full gap analysis for **multiple species** using GapAnalysis R package.
We provide a general approach using a for-loop for processing and storing results for multiple species. Example *Cucurbita* data provided within package library.

For each species, the workflow:

1. Runs SRSex on all records.

2. Checks inputs (occurrences, rasters, protected areas, ecoregions).

3. Runs gap analysis metrics.

4. Stores results.

5. **Creates an interactive leaflet map for the ERSex result, accessible as `ersex$map` for each species.**


***
#### Download GapAnalysis repository (*OPTIONAL*)
Copy the GapAnalysis GitHub repository locally as an alternative method to source functions and load example data **if GapAnalysis R package does not install properly in workflow below**. See [GitHub resource page](https://docs.github.com/en/get-started/start-your-journey/downloading-files-from-github) on copying a repository.

Easiest method: Download and extract the following [GapAnalysis repository](https://github.com/CIAT-DAPA/GapAnalysis/tree/master). Select green "Code" button on main page to download ZIP.

Can also clone the repository (if you have a GitHub account).

***
## Install GapAnalysis R package
GapAnalysis package is not currently available on CRAN. Install package via GitHub (if not installed already).

- If prompted to restart R, select YES.
- If prompted to update CRAN packages, input 1= ALL.
- If prompted "Do you want to install/source package that needs compilation?", select YES.

```{r load-helper-package, eval=FALSE}
# Load installation helper package
library(devtools)
```

```{r install-GapAnalysis-R-package, eval=FALSE}
# Install GapAnalysis using remotes from devtools
remotes::install_github("CIAT-DAPA/GapAnalysis")
```


***


## Data preparation
If you downloaded repository locally, your specific path to source functions and load data will differ based on your current working directly and the relative path to the downloaded repository.


```{r load-packages-and-data}
# Load packages
library(dplyr)
library(terra)
library(leaflet)
library(GapAnalysis)

# Load example data provided by the package
data("CucurbitaData")      # Occurrence data
data("CucurbitaRasts")     # Raster list
data("ProtectedAreas")     # Protected areas raster
data("ecoregions")         # Ecoregions data

# Example species (3): 
# 1. Cucurbita_cordata
# 2. Cucurbita_digitata
# 3. Cucurbita_palmata

# Prepare ecoregion and protected area layers
ecos <- terra::vect(ecoregions)
proAreas <- terra::unwrap(ProtectedAreas)

# Standardize the CRS across all spatial layers to prevent mismatch warnings
target_crs <- "+proj=longlat +datum=WGS84"
ecos <- terra::project(ecos, target_crs)
proAreas <- terra::project(proAreas, target_crs)

# Get unique species names and SDM rasters as a list
taxa <- unique(CucurbitaData$species)
sdms <- terra::unwrap(CucurbitaRasts)
sdmList <- lapply(seq_along(taxa), function(i) {
  r <- sdms[[i]]
  # Compare CRS as character strings
  if (terra::crs(r, proj=TRUE) != terra::crs(ecos, proj=TRUE)) {
    r <- terra::project(r, terra::crs(ecos))
  }
  # Convert 0 to NA, keeping only 1 and NA in the dataset
  r <- terra::subst(r, 0, NA)
  r[!(is.na(r) | r == 1)] <- NA
  r
})
names(sdmList) <- taxa
```


***

## Gap analysis for multiple species example usage
Example gap analysis workflow using a named list and for-loop (with raster value check) to process and store results.

```{r run-gap-analysis}
results_list <- list()

for (i in seq_along(taxa)) {
  taxon <- taxa[i]
  occurrenceData <- CucurbitaData[CucurbitaData$species == taxon, ]
  sdm <- sdms[[i]]
  if (terra::crs(sdm, proj=TRUE) != terra::crs(ecos, proj=TRUE)) {
    sdm <- terra::project(sdm, terra::crs(ecos))
  }
  sdm <- terra::subst(sdm, 0, NA)
  sdm[!(is.na(sdm) | sdm == 1)] <- NA

  # 1. Run SRSex first (on all records)
  srsex <- SRSex(taxon = taxon, occurrenceData = occurrenceData)

  # 2. Run check functions on inputs
  occurrences <- checkOccurrences(csv = occurrenceData, taxon = taxon)

  if (!inherits(sdm, "SpatRaster")) stop("sdm is not a SpatRaster")
  sdm_checked <- checksdm(sdm)
  proArea_checked <- checkProtectedAreas(protectedArea = proAreas, sdm = sdm_checked)
  
  # ---- PATCH: skip species if SDM raster has no values ----
  if (all(is.na(terra::values(sdm_checked)))) {
    message("Skipping ", taxon, ": SDM raster has no values.")
    next
  }

  # 3. Continue with main gap analysis
  gBuffer <- generateGBuffers(
    taxon = taxon,
    occurrenceData = occurrences$data,
    bufferDistM = 50000
  )

  # Ex-situ analysis
  grsex <- GRSex(
    taxon = taxon,
    sdm = sdm_checked,
    gBuffer = gBuffer
  )
  ersex <- ERSex(
    taxon = taxon,
    sdm = sdm_checked,
    occurrenceData = occurrences$data,
    gBuffer = gBuffer,
    ecoregions = ecos,
    idColumn = "ECO_NAME"
  )
  fcsex <- FCSex(
    taxon = taxon,
    srsex = srsex,
    grsex = grsex,
    ersex = ersex
  )

  # In-situ analysis
  srsin <- SRSin(
    taxon = taxon,
    sdm = sdm_checked,
    occurrenceData = occurrences$data,
    protectedAreas = proAreas
  )
  grsin <- GRSin(
    taxon = taxon,
    sdm = sdm_checked,
    protectedAreas = proAreas
  )
  ersin <- ERSin(
    taxon = taxon,
    sdm = sdm_checked,
    occurrenceData = occurrences$data,
    protectedAreas = proAreas,
    ecoregions = ecos,
    idColumn = "ECO_NAME"
  )
  fcsin <- FCSin(
    taxon = taxon,
    srsin = srsin,
    grsin = grsin,
    ersin = ersin
  )

  # Final combined FCS score
  fcsmean <- FCSc_mean(
    taxon = taxon,
    fcsin = fcsin,
    fcsex = fcsex
  )

  # 4. Store results for this taxon in the named list
  results_list[[taxon]] <- list(
    srsex = srsex,
    occurrences = occurrences,
    sdm_checked = sdm_checked,
    proArea_checked = proArea_checked,
    eco_checked = ecoregions,
    grsex = grsex,
    ersex = ersex,     # 5. Map created and stored
    fcsex = fcsex,
    srsin = srsin,
    grsin = grsin,
    ersin = ersin,
    fcsin = fcsin,
    fcsmean = fcsmean
  )
}
```

***
## Visualizing gap analysis results - Summary metrics tables

Results of gap analysis are organized as a selectable list, allowing for review of individual metrics. The tables below summarize the main metrics for the example species.

```{r results-tables}
# Ex-situ metrics
exsitu_table <- do.call(rbind, lapply(names(results_list), function(taxon) {
  r <- results_list[[taxon]]
  data.frame(
    Species          = taxon,
    SRS_exsitu       = r$srsex$`SRS exsitu`,
    GRS_exsitu       = r$fcsex$`GRS exsitu`,
    ERS_exsitu       = r$fcsex$`ERS exsitu`,
    FCS_exsitu       = r$fcsex$`FCS exsitu`,
    FCS_exsitu_score = r$fcsex$`FCS existu score`
  )
}))

# In-situ metrics
insitu_table <- do.call(rbind, lapply(names(results_list), function(taxon) {
  r <- results_list[[taxon]]
  data.frame(
    Species          = taxon,
    SRS_insitu       = r$fcsin$`SRS insitu`,
    GRS_insitu       = r$fcsin$`GRS insitu`,
    ERS_insitu       = r$fcsin$`ERS insitu`,
    FCS_insitu       = r$fcsin$`FCS insitu`,
    FCS_insitu_score = r$fcsin$`FCS insitu score`
  )
}))

# Combined metrics
combined_table <- do.call(rbind, lapply(names(results_list), function(taxon) {
  r <- results_list[[taxon]]
  data.frame(
    Species         = taxon,
    FCSc_min        = r$fcsmean$FCSc_min,
    FCSc_max        = r$fcsmean$FCSc_max,
    FCSc_mean       = r$fcsmean$FCSc_mean,
    FCSc_min_class  = r$fcsmean$FCSc_min_class,
    FCSc_max_class  = r$fcsmean$FCSc_max_class,
    FCSc_mean_class = r$fcsmean$FCSc_mean_class
  )
}))

knitr::kable(exsitu_table,   digits = 2, caption = "Ex-situ Gap Analysis metrics")
knitr::kable(insitu_table,   digits = 2, caption = "In-situ Gap Analysis metrics")
knitr::kable(combined_table, digits = 2, caption = "Combined Gap Analysis metrics")
```


***

## Visualizing gap analysis results - Maps

Each `ersex` result contains an **interactive leaflet map** in `ersex$map`.  
To display the map for each species in an R Markdown document, print the `$ersex$map` object for the species of interest.  
Examples provided below to show maps for example species:

```{r map-cucurbita-cordata}
# Show the ERSex interactive map for the first species (if available)
# Example species 1: Cucurbita cordata
if (!is.null(results_list[[1]])) results_list[[1]]$ersex$map
```


```{r map-cucurbita-digitata}
# Show the ERSex interactive map for the second species (if available)
# Example species 2: Cucurbita digitata
if (!is.null(results_list[[2]])) results_list[[2]]$ersex$map
```


```{r map-cucurbita-palmata}
# Show the ERSex interactive map for the third species (if available)
# Example species 3: Cucurbita palmata
if (!is.null(results_list[[3]])) results_list[[3]]$ersex$map
```

***
